Supplemental Online Materials

### 2023-04-24 Documents/manuscript.qmd
switch to Quarto

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
keywords: collective action, double standards, social identity theory, system justification theory, Black Lives Matter
abstract: "Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. In five studies (total *N* = 2,979), we investigated double standards in judging collective action---that is, whether observers will judge the same protest actions as more acceptable if the protesters are ingroup members and their cause aligns with the ingroup's interests (*identity-based double standards*) or if the protesters' cause aligns with the observers' ideological positions (*ideology-based double standards*). In two studies, we developed an instrument of 25 controversial protest actions, based in item response theory, to measure where people draw the line between acceptable and unacceptable forms of collective action. In three preregistered experiments, we used this instrument to test for double standards in judging collective action for workers' rights in the United Kingdom and for or against defunding the police and for or against restricting abortion in the United States. We found evidence for ideology-based, but not for identity-based, double standards: Participants judged the same protest actions to be more acceptable when the protesters' cause aligned with their own ideological position---but showed no consistent ingroup bias when judging collective action by ingroup and outgroup members. Our findings have theoretical and practical implications for understanding the often divided response to prominent social movements."
bibliography: references.bib
csl: apa.csl
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: yes
    template: apa-template.tex

---

```{r setup, include = FALSE}

@@

  # knitr options
  knitr::opts_chunk$set(echo = FALSE)

@@

In this paper, we examine double standards in judging collective action---that is, whether observers judge the same protest actions as more or less acceptable depending on their own and the protesters' group memberships (*identity-based double standards*) or on the protesters' cause and how it aligns with their own ideological positions (*ideology-based double standards*). In two studies, we first develop an instrument of 25 controversial protest actions, based in item response theory, to capture double standards in judging collective action. In three preregistered experiments, we then use this instrument to test for double standards in judging protest actions for workers' rights in the United Kingdom and for or against defunding the police and for or against restricting abortion in the United States. We find evidence for ideology-based, but not for identity-based, double standards: People judge the same protest actions to be more acceptable when the protesters' cause aligns with their own ideological positions---but do not show consistent ingroup bias when judging collective action by ingroup and outgroup members.

@@

Collective action---that is, any action taken by group members to advance a shared political goal [for similar definitions, see @becker_virtual_2012; @van_zomeren_building_2016]---is a powerful force driving social change. For example, Black Lives Matter protests lastingly shifted public discourse about racial inequity toward antiracist ideas [@dunivin_black_2022].

@@

In this research, we move this distinction from the researcher's intuition into the realm of scientific investigation and test for double standards in where people draw the line between acceptable (normative) and unacceptable (non-normative) forms of collective action. For that purpose, we define a double standard as judging the same action as more or less acceptable depending on who is performing the action and for what reason [for an analogous definition, see @foschi_double_2000].^[@feinberg_activists_2020 provided incidental evidence for double standards in judging collective action as, in a manipulation check, Black participants rated Black Lives Matter protests as less extreme than White participants and liberal participants rated protest actions for progressive causes as less extreme than conservative participants.] In this section, we derive predictions from social psychological theories about how identity-based and ideology-based double standards could lead people to judge the same protest actions as more or less acceptable depending on who the protesters are and what they are protesting.

@@

In the same vain, we propose that ingroup bias results in identity-based double standards in judging collective action by ingroup and outgroup members. That is, we hypothesize that people will judge the same protest actions as more acceptable if the protesters are ingroup members (*Hypothesis 1a*) or if the cause of the protest aligns with the ingroup's interests (*Hypothesis 1b*). This hypothesis could explain why, for example, far more Black than White Americans considered it appropriate for Black athletes to kneel down to protest police violence against Black Americans [@yougov_nfl_2017]. Notably, this hypothesis applies equally to historically disadvantaged groups mobilizing against social injustice and to historically advantaged groups defending their group's position.

@@

As system-justifying beliefs are closely related to political conservatism [@jost_political_2003], this hypothesis could explain partisan differences in what actions liberals and conservatives consider acceptable means of protest for progressive and conservative causes. Ideology-based double standards could, for example, explain why far more Democrats than Republicans considered it appropriate to kneel during the national anthem to protest racist police violence [@yougov_nfl_2017].

@@

So far, we ...

<!-- Throughout this paper, we will use "identity-based" as a shorthand for double standards based on social categories such as race, gender, and class and "ideology-based" as a shorthand for double standards based on system-justification and political ideology. -->

@@

We included additional measures to describe the sample, to describe reactions to the manipulation, or to use in non-preregistered analyses. In addition to demographic questions, we asked participants how outraged they were about recent incidents of police violence against Black Americans, to what extent they identifed with the protesters described in the study, and to what extent they identifed with their racial ingroup (1 = *not at all*, 7 = *very much*). We also asked how often, if at all, participants had participated in protest actions such as the ones we had asked about (1 = *never*, 5 = *very often*) and whether they has participated in protests for reforming, defunding or abolishing the police; against reforming, defunding or abolishing the police; or in neither. We measured political orientation with a one-item semantic differential scale: "People often describe their political orientation as liberal or conservative. On a scale from liberal to conservative, where would you position yourself?" (1 = *liberal*, 7 = *conservative*).

@@

We preregistered the sample size, hypotheses, inclusion/exclusion criteria, measures, and manipulations ([https://osf.io/kphyw/?view_only=59f5792a6d3f4fa48516ea9d5f637822](https://osf.io/kphyw/?view_only=59f5792a6d3f4fa48516ea9d5f637822)). We deviated from the preregistration in two ways. First, we stopped collecting data before reaching the preregistered sample size. We did so as the Supreme Court decision on reproductive rights, announced on June 24, 2022, rendered the experimental manipulation invalid. Second, we do not report results for two further hypotheses---testing moral conviction as a potential mechanism underlying ideology-based double standards---as those findings are beyond the scope of this manuscript and will be reported in a future publication. We made all materials, data, and analysis scripts available online ([https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62](https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62)).

@@

Growing evidence shows that, in many situations, observers apply a more lenient standards when judging moral transgressions by ingroup members [@abrams_double_2013; @endevelt_everyone_2021]. And yet, our findings suggest that, unlike judgments in other domains, judgments about collective action are not subject to double standards based on group identities such as class, race, or gender. Why did we not find any evidence for identity-based double standards?

@@

Our findings provided a relevant comparison to findings from a recent experiment in which White Americans viewed pictures of either peaceful or destructive Black Lives Matter protests and rated how legitimate (i.e., fair, reasonable, legitimate, and justified) they considered the pictured protests to be [@teixeira_white_2022]. In this experiment, more conservative participants rated destructive and, to a lesser extent, peaceful protests to be less legitimate than more liberal participants. Likewise, participants who identified more strongly with their ethnic ingroup rated destructive and, to a lesser extent, peaceful protests to be less legitimate than participants who identified less strongly with their ingroup. However, participants' group identification was no longer associated with their protest ratings after controlling for their political orientation.^[These findings are based on our reanalysis, reported in the SOM, of Teixeira et al.'s [-@teixeira_white_2022] experiment. We thank the authors for making their data available online.] In this way, results from the experiment aligned with our conclusion that political ideologies are more important than social identities for shaping how people judge controversial forms of collective action.

@@

<div id="refs"></div>

>>>>

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
abstract: "Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. In five studies (total *N* = 2,979), we investigated double standards in judging collective action---that is, whether observers will judge the same protest actions as more acceptable if the protesters are ingroup members and their cause aligns with the ingroup's interests (*identity-based double standards*) or if the protesters' cause aligns with the observers' ideological positions (*ideology-based double standards*). In two studies, we developed an instrument of 25 controversial protest actions, based in item response theory, to measure where people draw the line between acceptable and unacceptable forms of collective action. In three preregistered experiments, we used this instrument to test for double standards in judging collective action for workers' rights in the United Kingdom and for or against defunding the police and for or against restricting abortion in the United States. We found evidence for ideology-based, but not for identity-based, double standards: Participants judged the same protest actions to be more acceptable when the protesters' cause aligned with their own ideological position---but showed no consistent ingroup bias when judging collective action by ingroup and outgroup members. Our findings have theoretical and practical implications for understanding the often divided response to prominent social movements."
keywords: collective action, double standards, social identity theory, system justification theory, Black Lives Matter
bibliography: references.bib
csl: apa.csl
execute:
  echo: false
format:
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    template: apa-template.tex
---

```{r}
#| include: false

@@

@@

In this paper, we examine double standards in judging collective action---that is, whether observers judge the same protest actions as more or less acceptable depending on their own and the protesters' group memberships (*identity-based double standards*) or on the protesters' cause and how it aligns with their own ideological positions (*ideology-based double standards*). In two studies, we first develop an instrument of 25 controversial protest actions, based in item response theory, to capture double standards in judging collective action. In three preregistered experiments, we then use this instrument to test for double standards in judging protest actions for workers' rights in the United Kingdom and for or against defunding the police and for or against restricting abortion in the United States. We find evidence for ideology-based, but not for identity-based, double standards: People judge the same protest actions to be more acceptable when the protesters' cause aligns with their own ideological positions---but do not show consistent ingroup bias (in terms of class, race, and gender) when judging collective action by ingroup and outgroup members.

@@

Collective action---that is, any action taken by group members to advance a shared political goal [for similar definitions, see @becker_virtual_2012; @van_zomeren_building_2016]---is a powerful force driving social change. For example, Black Lives Matter protests shifted public discourse about racial inequity toward antiracist ideas [@dunivin_black_2022].

@@

In this research, we move this distinction from the researcher's intuition into the realm of scientific investigation and test for double standards in where people draw the line between acceptable (normative) and unacceptable (non-normative) forms of collective action. For that purpose, we define a double standard as judging the same protest actions as more or less acceptable depending on who the protesters are and what they are protesting. [for an analogous definition, see @foschi_double_2000].^[@feinberg_activists_2020 provided incidental evidence for double standards in judging collective action as, in a manipulation check, Black participants rated Black Lives Matter protests as less extreme than White participants and liberal participants rated protest actions for progressive causes as less extreme than conservative participants.]

@@

In the same vein, we propose that ingroup bias results in identity-based double standards in judging collective action by ingroup and outgroup members. That is, we hypothesize that people will judge the same protest actions as more acceptable if the protesters are ingroup members (*Hypothesis 1a*) or if the cause of the protest aligns with the ingroup's interests (*Hypothesis 1b*). This hypothesis could explain why, for example, far more Black than White Americans considered it appropriate for Black athletes to kneel down to protest police violence against Black Americans [@yougov_nfl_2017]. Notably, this hypothesis applies equally to historically disadvantaged groups mobilizing against social injustice and to historically advantaged groups defending their group's position.

@@

As system-justifying beliefs are closely related to political conservatism [@jost_political_2003], this hypothesis could explain partisan differences in what actions liberals and conservatives consider acceptable means of protest for progressive and conservative causes---for example, why far more Democrats than Republicans considered it appropriate to kneel during the national anthem to protest racist police violence [@yougov_nfl_2017].

@@

<!-- Write it in your own words and structure it (with references) later. -->
<!-- Political orientation can be understood as ideology and identity. -->

- Are Partisan Double Standards Based in Ideology or Identity?
- Evidence for partisanship as a elective affinity [@jost_left_2021]
- Evidence for partisanship as a social identity
- We will use "identity-based" as a shorthand for double standards based on assigned categories such as class, race, and gender and "ideology-based" as a shorthand for double standards based on system-justification and political orientation.

<!-- I could move the last paragraph of the previous section to here. -->

@@

We included additional measures to describe the sample, to describe reactions to the manipulation, or to use in non-preregistered analyses. In addition to demographic questions, we asked participants how outraged they were about recent incidents of police violence against Black Americans, to what extent they identifed with the protesters described in the study, and to what extent they identifed with their racial ingroup (1 = *not at all*, 7 = *very much*). We also asked how often, if at all, participants had participated in protest actions such as the ones we had asked about (1 = *never*, 5 = *very often*) and whether they had participated in protests for reforming, defunding or abolishing the police; against reforming, defunding or abolishing the police; or in neither. We measured political orientation with a one-item semantic differential scale: "People often describe their political orientation as liberal or conservative. On a scale from liberal to conservative, where would you position yourself?" (1 = *liberal*, 7 = *conservative*).

@@

We preregistered the sample size, hypotheses, inclusion/exclusion criteria, measures, and manipulations ([https://osf.io/kphyw/?view_only=59f5792a6d3f4fa48516ea9d5f637822](https://osf.io/kphyw/?view_only=59f5792a6d3f4fa48516ea9d5f637822)). We deviated from the preregistration in two ways. First, we stopped collecting data before reaching the preregistered sample size. We did so as the Supreme Court decision on reproductive rights, announced on June 24, 2022, rendered the experimental manipulation invalid. Second, we do not report results for two further hypotheses---testing moral conviction as a potential mechanism underlying ideology-based double standards---as those findings are beyond the scope of this manuscript and will be reported in a future publication. We made all materials, data, and analysis scripts available online ([https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62](https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62)). Our sample size was determined by budget and time constraints.

@@

Growing evidence shows that, in many situations, observers apply a more lenient standards when judging moral transgressions by ingroup members [@abrams_double_2013; @endevelt_everyone_2021]. And yet, our findings suggest that, unlike judgments in other domains, judgments about collective action are not subject to double standards based on group identities such as class, race, or gender. 
<!-- Why did we not find any evidence for identity-based double standards? -->

@@

<!-- Our findings provided a relevant comparison to findings from a recent experiment in which White Americans viewed pictures of either peaceful or destructive Black Lives Matter protests and rated how legitimate (i.e., fair, reasonable, legitimate, and justified) they considered the pictured protests to be [@teixeira_white_2022]. In this experiment, more conservative participants rated destructive and, to a lesser extent, peaceful protests to be less legitimate than more liberal participants. Likewise, participants who identified more strongly with their ethnic ingroup rated destructive and, to a lesser extent, peaceful protests to be less legitimate than participants who identified less strongly with their ingroup. However, participants' group identification was no longer associated with their protest ratings after controlling for their political orientation.^[These findings are based on our reanalysis, reported in the SOM, of Teixeira et al.'s [-@teixeira_white_2022] experiment. We thank the authors for making their data available online.] In this way, results from the experiment aligned with our conclusion that political ideologies are more important than social identities for shaping how people judge controversial forms of collective action. -->

@@

::: {#refs}
:::

### 2023-04-24 Documents/supplemental-materials.qmd
switch to Quarto

---
title: Supplemental Online Materials
subtitle: Double Standards in Judging Collective Action
bibliography: references.bib
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: yes
    template: som-template.tex
csl: apa.csl
---

@@

>>>>

---
title: Supplemental Online Materials
subtitle: Double Standards in Judging Collective Action
bibliography: references.bib
csl: apa.csl
execute:
  echo: false
format:
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    template: som-template.tex
---

@@

```{=latex}
\setcounter{table}{3}

\begin{table}[t!]
\caption{Correlations between variables in Experiment 1}
\centering
\begin{tabular}{rlrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 \\ \midrule
 1 & Proportion            &      & -.31 & -.16 & -.23 \\
 2 & Political Orientation & -.31 &      &  .41 &  .41 \\
 3 & Social Dominance      & -.16 &  .41 &      &  .29 \\
 4 & System Justification  & -.23 &  .41 &  .29 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); Political Orientation (1 = \textit{left}, 7 = \textit{right}); Social Dominance Orientation (\textit{factor score}); Systen-Justifying Beliefs (\textit{factor score})}
\end{table}
```

```{=latex}
\setcounter{table}{4}

\begin{table}[t!]
\caption{Correlations between variables in Experiment 2}
\centering
\begin{tabular}{rlrrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 &    5 \\ \midrule
 1 & Proportion            &      & -.21 &  .29 & -.23 & -.07 \\
 2 & System Justification  & -.21 &      & -.45 &  .47 &  .04 \\
 3 & Policy Support        &  .29 & -.45 &      & -.49 &  .00 \\
 4 & Political Orientation & -.23 &  .47 & -.49 &      &  .03 \\
 5 & Identification        & -.07 &  .04 &  .00 &  .03 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); Systen-Justifying Beliefs (\textit{factor score}); Policy Support (1 = \textit{strongly oppose}, 5 = \textit{strongly support}); Political Orientation (1 = \textit{liberal}, 7 = \textit{conservative}); Racial Identification (1 = \textit{not at all}, 7 = \textit{very much})}
\end{table}
```

```{=latex}
\setcounter{table}{5}

\begin{table}[t!]
\caption{Correlations between variables in Experiment 3}
\centering
\begin{tabular}{rlrrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 &    5 \\ \midrule
 1 & Proportion            &      & -.27 & -.22 & -.32 & -.11 \\
 2 & System Justification  & -.27 &      &  .61 &  .78 &  .16 \\
 3 & Policy Support        & -.22 &  .61 &      &  .74 &  .16 \\
 4 & Political Orientation & -.32 &  .78 &  .74 &      &  .20 \\
 5 & Identification        & -.11 &  .16 &  .16 &  .20 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); Systen-Justifying Beliefs (\textit{factor score}); Policy Support (1 = \textit{strongly oppose}, 5 = \textit{strongly support}); Political Orientation (1 = \textit{liberal}, 7 = \textit{conservative}); Gender Identification (1 = \textit{not at all}, 7 = \textit{very much})}
\end{table}
```
